```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figures/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

# Loading and preprocessing the data

First, lets load our data
```{r load_data, echo=TRUE}
activity = read.csv("activity.csv")
```

and see its summary
```{r str_data, echo=TRUE}
str(activity)
```
We see, that the second column has factor type. That is not what we need, so let's transform its type to date.
```{r factor_to_date, echo=TRUE}
activity$date <- as.Date(activity$date)
```
Let's check, that our transformation has effect:
```{r chacking_factor_to_date_coercion, echo=TRUE}
str(activity)
```

Now let's see, what columns have NAs:
```{r detect_columns_with_na, echo=TRUE}
sum(is.na(activity$steps))
sum(is.na(activity$date))
sum(is.na(activity$interval))
```
So, only **steps** column has NA values.

We will also need **dplyr** library, so let' load it:
```{r load_dplyr, echo=TRUE}
library("dplyr")
```

# What is mean total number of steps taken per day?

First, we will group **activity** table by days
```{r activity_grouped_by_days, echo=TRUE}
days <- group_by(activity, date)
```
and then calculate total number of steps, that were made each day (removing NA values)
```{r total_number_of_steps, echo=TRUE}
total_steps <- summarise(days, steps=sum(steps, na.rm=TRUE))
head(total_steps)
```

Now we will make a histogram:
```{r total_number_of_steps_hist, echo=TRUE}
hist(total_steps$steps, col="green", breaks=30, xlab="Steps", main="Total number of steps")
```


The mean number of steps is:
```{r mean_number_of_steps, echo=TRUE}
mean(total_steps$steps)
```

The median number:
```{r median_number_of_steps, echo=TRUE}
median(total_steps$steps)
```

# What is the average daily activity pattern?

Let's group our data by five-minute intervals
```{r group_by_intervals, echo=TRUE}
intervals <- group_by(activity, interval)
```
and the calculate average number of steps in each interval:
```{r average_number_of_steps, echo=TRUE}
average_steps_num <- summarise(intervals, avg=mean(steps, na.rm=TRUE))
```

```{r average_number_of_steps_plot, echo=TRUE}
with(average_steps_num, 
     plot(avg, type="l", col="red", lwd=3, xlab="interval", ylab="steps", main="Average # of steps by interval"))
```

The interval, that contains average maximum is:
```{r average_max_interval, echo=TRUE}
which.max(average_steps_num$avg)
```

# Imputing missing values

The number of missing values is:
```{r na_number, echo=TRUE}
sum(is.na(activity$steps))
```

We will replace missing values with the mean of other values in the same interval:
```{r imputing_values, echo=TRUE}
average_steps_num$avg <- round(average_steps_num$avg)
imputed <- merge(activity, average_steps_num, by.x = "interval", by.y="interval")
imputed <- arrange(imputed, date)
imputed$steps <- ifelse(is.na(imputed$steps), imputed$avg, imputed$steps)
```

Grouping by days:
```{r activity_grouped_by_days_imputed, echo=TRUE}
days_imputed <- group_by(imputed, date)
```
and then calculate total number of steps, that were made each day (removing NA values)
```{r total_number_of_steps_imputed, echo=TRUE}
total_steps_imputed <- summarise(days_imputed, steps=sum(steps, na.rm=TRUE))
head(total_steps_imputed)
```
Now we will make a histogram:
```{r total_number_of_steps_hist_imputed, echo=TRUE}
hist(total_steps_imputed$steps, col="green", breaks=30, xlab="Steps", main="Total number of steps")
```


The mean number of steps is:
```{r mean_number_of_steps_imputed, echo=TRUE}
mean(total_steps_imputed$steps)
```

The median number:
```{r median_number_of_steps_imputed, echo=TRUE}
median(total_steps_imputed$steps)
```

# Are there differences in activity patterns between weekdays and weekends?

We will add a factor column:
```{r creating_factor_column, echo=TRUE}
imputed$wday <- ifelse(weekdays(imputed$date)=='Суббота' | weekdays(imputed$date)=='Воскресенье', 
                       'Weekend', "Workday")
imputed$wday <- as.factor(imputed$wday)
```

Let's see if it works:
```{r factor_column_testing, echo=TRUE}
head(imputed)
```

Now we will get average number of steps for each interval of a workday:
```{r select_workdays, echo=TRUE}
workdays <- filter(imputed, wday=='Workday')  %>% 
            group_by(interval) %>% summarise( avg=mean(steps), wday='Workday')
```

...and of a weekend

```{r select_weekends, echo=TRUE}
weekends <- filter(imputed, wday=='Weekend') %>% 
            group_by(interval) %>% summarise( avg=mean(steps), wday='Weekend')
```

```{r merge_workdays_and_weekends, echo=TRUE}
by_day_type <- rbind(workdays, weekends)
by_day_type$wday <- as.factor(by_day_type$wday)
head(by_day_type)
```


```{r average_number_of_steps_plot_by_wday, echo=TRUE}
     library(lattice)
     xyplot(avg ~ interval | wday, data=by_day_type, layout=c(1,2), type="l")
```